# Platform Correlations

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(pheatmap)
library(mcr)
library(furrr)
library(progressr)
```

```{r}
# Olink -----
olink_b1 <- read_tsv("data/data_phase2_batch1_curated_20241217.tsv")
olink_b4 <- read_tsv("data/data_phase2_batch4_curated_20250425.tsv")
olink_b5 <- read_tsv("data/data_phase2_batch5_curated_20250701.tsv")
common_olink_cols <- intersect(intersect(colnames(olink_b1), colnames(olink_b4)), colnames(olink_b5))
common_olink_assays <- intersect(
  intersect(
    unique(olink_b1 |> pull(Assay)), 
    unique(olink_b4 |> pull(Assay))
  ), 
  unique(olink_b5 |> pull(Assay))
)

olink <- bind_rows(
  olink_b1 |> select(all_of(common_olink_cols)) |> filter(Assay %in% common_olink_assays) |> mutate(Batch = 1), 
  olink_b4 |> select(all_of(common_olink_cols)) |> filter(Assay %in% common_olink_assays) |> mutate(Batch = 4),
  olink_b5 |> select(all_of(common_olink_cols)) |> filter(Assay %in% common_olink_assays) |> mutate(Batch = 5)
  ) |>
    select("DAid", "Assay", "UniProt", "Batch", "NPX", "LOD")

rm(olink_b1)
rm(olink_b4)
rm(olink_b5)

# Soma -----
soma_wide <- read_tsv("data/data_phase2_somalogic_curated_20250922.tsv")
soma_meta <- read_tsv("data/meta_phase2_somalogic_20250922.tsv")

soma <- soma_wide |>
  pivot_longer(cols = -c("DAid"), names_to = "AptName", values_to = "Value") |>
  left_join(soma_meta |> filter(Organism == "Human") |> select(AptName, EntrezGeneSymbol), by = "AptName") |>
  rename(Assay = EntrezGeneSymbol)

rm(soma_wide)
rm(soma_meta)

sample_meta <- readxl::read_excel("data/samples_2025-10-10.xlsx") |>
  select(DAid, Disease, Age, Sex, BMI)

gene_meta <- read_csv("data/gene_metadata.csv")

protein_tiers <- readxl::read_excel("data/proteins_classified_tiers.xlsx") |>
  select(Assay = Gene, Tier = `confidence tier`) |>
  mutate(Tier = as.factor(Tier))
```

```{r}
pal_secreted <- c(
  "Actively secreted to blood" = "#B24F52",
  "Secreted in female reproductive system" = "#2A9D8F",
  "Unknown secreted location" = "#7C7C7C",
  "Secreted in male reproductive system" = "#A9D18E",
  "Secreted to extracellular matrix" = "#A6CEE3",
  "Secreted to digestive system" = "#264653",
  "Secreted in brain" = "#F4A5AE",
  "Secreted to other locations" = "#B89B74",
  "Digestive system" = "#264653",
  "Intracellular and Membrane" = "#457B9D",
  "Not part of the secretome annotation" = "lightgrey"
)
```

## Global correlation of common proteins

Select common proteins and DAids.

```{r}
common_daids <- intersect(unique(olink |> pull(DAid)), unique(soma |> pull(DAid)))
common_assays <- intersect(unique(olink |> pull(Assay)), unique(soma |> pull(Assay)))

olink <- olink |> filter(DAid %in% common_daids, Assay %in% common_assays)
soma <- soma |> filter(DAid %in% common_daids, Assay %in% common_assays)

meta <- sample_meta |> filter(DAid %in% common_daids)
```

Aptamers per assay.

```{r}
apt_per_assay <- soma |>
  group_by(Assay) |>
  summarise(n_apt = n_distinct(AptName))

ggplot(apt_per_assay, aes(x = n_apt)) +
  stat_bin(binwidth = 1, fill = "lightblue", alpha = 0.7) + 
  labs(x = "Unique AptName Count", y = "Count") +
  theme_bw()
```

Spearman correlations. 

```{r}
# Calculate the Spearman correlation for each common Assay
correlation <- olink |>
  inner_join(soma, by = c("DAid", "Assay")) |>
  group_by(Assay, AptName) |>
  summarise(correlation = cor.test(NPX, Value, method = "spearman")$estimate) |>
  ungroup() |>
  left_join(gene_meta, by = "Assay") |>
  mutate(secretome_annotation = replace_na(secretome_annotation, "Unknown secreted location"))

median_correlation <- correlation |>
  pull(correlation) |>
  median()

# Plot a histogram of the correlation coefficient
ggplot(correlation, aes(x = correlation, fill = secretome_annotation)) +
  geom_histogram(binwidth = 0.1, color = "black", alpha = 0.7, position = "stack") +
  geom_text(x = median_correlation + 0.2, y = 1000, label = round(median_correlation, 4), color = "black", size = 5) +
  geom_vline(aes(xintercept = median_correlation), linetype = "dashed", color = "black") +
  labs(x = "Correlation Coefficient", y = "Proportion", fill = "Group") +
  scale_fill_manual(values = pal_secreted) + 
  theme_bw()

ggsave("results/platform_corr_global.png", width = 8, height = 4)
```

```{r}
# Calculate the Spearman correlation for each common Assay for each Olink batch
# correlation <- olink |>
#   inner_join(soma, by = c("DAid", "Assay")) |>
#   group_by(Assay, AptName, Batch) |>
#   summarise(correlation = cor.test(NPX, Value, method = "spearman")$estimate) |>
#   dplyr::ungroup()

# median_correlation <- correlation |>
#   pull(correlation) |>
#   median()

# # Plot a histogram of the correlation coefficient
# ggplot(correlation, aes(x = correlation)) +
#   stat_bin(binwidth = 0.1, fill = "lightblue", alpha = 0.7) + 
#   geom_vline(aes(xintercept = median_correlation), linetype = "dashed", color = "black") +
#   geom_text(x = median_correlation + 0.2, y = 2500, label = round(median_correlation, 4), color = "black", size = 5) +
#   labs(x = "Correlation Coefficient", y = "Count") +
#   theme_bw()

# ggsave("results/platform_corr_global_perbatch.png", width = 4, height = 3)
```

Per samples correlation

```{r}
# Join the olink and soma data
merged_dat <- soma |>
  left_join(olink, by = c("DAid", "Assay"))

# Calculate the correlation for each sample
sample_correlation <- merged_dat |> 
  group_by(DAid) |> 
  summarise(correlation = cor.test(NPX, Value, method = "spearman")$estimate) |> 
  ungroup()

# Calculate the median correlation per sample
median_correlation <- sample_correlation |> 
  pull(correlation) |> 
  median()

# Plot a histogram of the correlation coefficient per sample
ggplot(sample_correlation, aes(x = correlation)) +
  stat_bin(binwidth = 0.05, fill = "lightblue", alpha = 0.7) + 
  geom_vline(aes(xintercept = median_correlation), linetype = "dashed", color = "black") +
  geom_text(x = median_correlation + 0.2, y = 1000, label = round(median_correlation, 4), color = "black", size = 5) +
  labs(x = "Correlation Coefficient", y = "Count") +
  theme_bw()

ggsave("results/platform_corr_global_per_sample.png", width = 4, height = 3)

# Merge the selected_dat and meta data frames by DAid
merged_dat <- merged_dat |> 
  left_join(meta, by = "DAid") |>
  left_join(sample_correlation, by = "DAid")

# Create a violin plot of correlation by disease
ggplot(merged_dat, aes(x = Disease, y = correlation)) +
  geom_violin(aes(fill = Disease), trim = F, alpha = 0.7) +
  geom_segment(aes(x = Disease, xend = Disease, y = median(correlation), yend = median(correlation)), colour = "black") +
  labs(x = "Disease", y = "Correlation") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

ggsave("results/platform_corr_global_violin_by_disease.png", width = 7, height = 4)
```

Correlation per Disease

```{r}
olink_meta <- olink |>
  inner_join(meta, by = "DAid")

# Compute per-disease correlation per protein
r_disease <- olink_meta |>
  inner_join(soma, by = c("DAid", "Assay")) |>
  group_by(Disease, Assay, AptName) |>
  summarise(
    r = if (sum(is.finite(NPX) & is.finite(Value)) > 1)
      suppressWarnings(cor(NPX, Value, method = "spearman", use = "complete.obs"))
    else NA_real_,
    .groups = "drop"
  )

# Summarise per-disease median correlation per protein
r_matrix <- r_disease |>
  group_by(Assay, AptName, Disease) |>
  summarise(median_r = median(r, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Disease, values_from = median_r)

# Clustered heatmap (proteins × diseases)
r_mat <- r_matrix |> select(where(is.numeric)) |> as.data.frame()
rownames(r_mat) <- paste(r_matrix$Assay, r_matrix$AptName, sep = "_")

row_ann <- r_matrix |>
  left_join(gene_meta, by = "Assay") |>
  mutate(secretome_annotation = replace_na(secretome_annotation, "Unknown secreted location")) |>
  select(Assay, AptName, secretome_annotation) |>
  mutate(rowname = paste(Assay, AptName, sep = "_")) |>
  tibble::column_to_rownames("rowname") |>
  select(secretome_annotation)

row_ann <- row_ann[rownames(r_mat), , drop = FALSE]
ann_colors <- list(secretome_annotation = pal_secreted)

heatmap <- pheatmap(
  r_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  na_col = "grey80",
  main = "Per-Disease Median Spearman Correlation",
  annotation_row = row_ann,
  annotation_colors = ann_colors
)

ggsave("results/platform_corr_global_per_disease.png", heatmap, width = 8, height = 18)

# Identify proteins with large overall disagreement
# Disagreement defined as high cross-disease variability in correlation
r_var <- r_disease |>
  group_by(Assay, AptName) |>
  summarise(sd_r = sd(r, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(sd_r))

# Flag proteins with large variance in per-disease correlation
flagged <- r_var |> filter(sd_r > 0.35)

# Plot per-disease correlations for flagged proteins
flagged_plot <- r_disease |>
  inner_join(flagged, by = c("Assay", "AptName")) |>
  mutate(Protein = paste(Assay, AptName, sep = "_"))

ggplot(flagged_plot, aes(x = Disease, y = r, group = Protein)) +
  geom_point() +
  geom_line(aes(color = Protein), alpha = 0.6) +
  theme_bw() +
  labs(
    x = "Disease group",
    y = "Spearman correlation (Olink vs Soma)",
    title = "Per-disease correlation for proteins with disease-dependent disagreement"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

ggsave("results/variable_per_disease_proteins_corr.png", width = 8, height = 4)
```

Correlations for different Olink protein LOD bins

```{r}
below_LOD <- olink |>
  group_by(Assay) |>
  summarise(
    prop_below_LOD = mean(NPX < LOD, na.rm = TRUE)
  )

bins <- c(0, 0.1, 0.25, 0.5, 0.75, 1)
below_LOD <- below_LOD |>
  mutate(bin = cut(prop_below_LOD, breaks = bins, include.lowest = TRUE))

correlation <- olink |>
  inner_join(soma, by = c("DAid", "Assay")) |>
  group_by(Assay, AptName) |>
  summarise(
    correlation = suppressWarnings(cor.test(NPX, Value, method = "spearman")$estimate),
    .groups = "drop"
  )

correlation_binned <- correlation |>
  left_join(below_LOD, by = "Assay") |>
  left_join(gene_meta, by = "Assay") |>
  mutate(secretome_annotation = replace_na(secretome_annotation, "Unknown secreted location")) |>
  mutate(
    secretome_annotation = if_else(
      secretome_annotation != "Actively secreted to blood", 
      "Non secreted in blood", 
      "Actively secreted to blood"
    )
  ) |>
  left_join(protein_tiers, by = "Assay")

ggplot(correlation_binned, aes(x = bin, y = correlation)) +
  geom_violin(trim = FALSE, color = "black") +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  theme_bw() +
  labs(
    x = "Proportion of samples below LOD (Olink)",
    y = "Correlation Coefficient"
  )

ggsave("results/platform_corr_global_bins.png", width = 10, height = 4)

ggplot(correlation_binned, aes(x = bin, y = correlation, fill = secretome_annotation)) +
  geom_violin(trim = FALSE, color = "black") +
  geom_boxplot(width = 0.1, outlier.shape = NA, position = position_dodge(width = 0.9)) +
  theme_bw() +
  scale_fill_manual(values = c(
    "Actively secreted to blood" = "#B24F52",
    "Non secreted in blood" = "#7C7C7C"
  )) +
  labs(
    x = "Proportion of samples below LOD (Olink)",
    y = "Correlation Coefficient"
  )

ggsave("results/platform_corr_global_bins_secretome.png", width = 10, height = 4)

ggplot(correlation_binned, aes(x = bin, y = correlation, fill = Tier)) +
  geom_violin(trim = FALSE, color = "black") +
  geom_boxplot(width = 0.1, outlier.shape = NA, position = position_dodge(width = 0.9)) +
  theme_bw() +
  scale_fill_brewer() + 
  labs(
    x = "Proportion of samples below LOD (Olink)",
    y = "Correlation Coefficient"
  )


ggsave("results/platform_corr_global_bins_tiers.png", width = 10, height = 4)
```

Correlations for Assays/Samples above LOD vs bellow LOD (Olink)

```{r}
# Separate data based on LOD
olink_below <- olink |> filter(NPX < LOD) |> mutate(LOD_group = "Below LOD")
olink_above <- olink |> filter(NPX >= LOD) |> mutate(LOD_group = "Above LOD")

olink_split <- bind_rows(olink_below, olink_above)

# Compute correlation per Assay and group (below/above LOD)
correlation <- olink_split |>
  inner_join(soma, by = c("DAid", "Assay")) |>
  group_by(LOD_group, Assay, AptName) |>
  summarise(
    correlation = if (sum(is.finite(NPX) & is.finite(Value)) > 1)
      suppressWarnings(cor(NPX, Value, method = "spearman", use = "complete.obs"))
    else
      NA_real_,
    .groups = "drop"
  ) |>
  filter(!is.na(correlation))

median_corr <- correlation |>
  group_by(LOD_group) |>
  summarise(median_correlation = median(correlation, na.rm = TRUE))

ggplot(correlation, aes(x = correlation)) +
  stat_bin(binwidth = 0.1, fill = "lightblue", alpha = 0.7) +
  geom_vline(
    data = median_corr,
    aes(xintercept = median_correlation),
    linetype = "dashed",
    color = "black"
  ) +
  geom_text(
    data = median_corr,
    aes(x = median_correlation + 0.4, y = Inf,
        label = round(median_correlation, 4)),
    vjust = 2, color = "black", size = 4
  ) +
  labs(x = "Spearman Correlation", y = "Count") +
  facet_wrap(~ LOD_group, nrow = 1) +
  theme_bw()

ggsave("results/platform_corr_global_LOD_split.png", width = 6, height = 3)

```

```{r}
# Compute correlation including all sample
r_all <- olink |>
  inner_join(soma, by = c("DAid", "Assay")) |>
  group_by(Assay, AptName) |>
  summarise(
    r_all = if (sum(is.finite(NPX) & is.finite(Value)) > 1)
      suppressWarnings(cor(NPX, Value, method = "spearman", use = "complete.obs"))
    else
      NA_real_,
    .groups = "drop"
  )

# Compute correlation for detected (above LOD) samples only
r_detected <- olink |>
  filter(NPX >= LOD) |>
  inner_join(soma, by = c("DAid", "Assay")) |>
  group_by(Assay, AptName) |>
  summarise(
    r_detected_only = if (sum(is.finite(NPX) & is.finite(Value)) > 1)
      suppressWarnings(cor(NPX, Value, method = "spearman", use = "complete.obs"))
    else
      NA_real_,
    .groups = "drop"
  )

comparison <- inner_join(r_all, r_detected, by = c("Assay", "AptName")) |>
  mutate(delta_r = r_detected_only - r_all) |>
  left_join(below_LOD, by = "Assay")

# Identify proteins where correlation collapses (large negative delta)
collapsed_proteins <- comparison |>
  filter(!is.na(r_all), !is.na(r_detected_only)) |>
  filter(r_all < 0.3 & r_detected_only > 0.7) |>
  arrange(r_all)

ggplot(comparison, aes(x = r_all, y = r_detected_only, color = bin)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    x = "Spearman correlation (all samples)",
    y = "Spearman correlation (samples above LOD)"
  ) +
  scale_colour_brewer(palette = "Set1") +
  theme_bw()

ggsave("results/platform_corr_global_scatter_detected_vs_all.png", width = 7, height = 5)
```

Soma intra-correlation vs correlation with Olink

```{r}
aptamer_counts <- apt_per_assay |>
  filter(n_apt > 1) 

soma_multi <- soma |>
  semi_join(aptamer_counts, by = "Assay")

intra_corr <- soma_multi |>
  group_by(Assay) |>
  summarise(
    cor_matrix = list({
      df <- pivot_wider(cur_data(), names_from = AptName, values_from = Value)
      mat <- cor(df |> select(-DAid), use = "pairwise.complete.obs", method = "spearman")
      mat[lower.tri(mat)]
    }),
    .groups = "drop"
  ) |>
  unnest_longer(cor_matrix, values_to = "r_intra")

aptamer_olink_corr <- soma_multi |>
  inner_join(olink, by = c("DAid", "Assay")) |>
  group_by(Assay, AptName) |>
  summarise(
    r_olink = if (sum(is.finite(Value) & is.finite(NPX)) > 1)
      suppressWarnings(cor(Value, NPX, method = "spearman", use = "complete.obs"))
    else NA_real_,
    .groups = "drop"
  )

# Compute median intra-aptamer r per protein
median_intra <- intra_corr |>
  group_by(Assay) |>
  summarise(median_intra = median(r_intra, na.rm = TRUE))

# Combine with aptamer-specific Olink correlations
combined <- aptamer_olink_corr |>
  left_join(median_intra, by = "Assay") |>
  left_join(below_LOD, by = "Assay")

ggplot(combined, aes(x = median_intra, y = r_olink, color = bin)) +
  geom_point(size = 1) +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  labs(
    x = "Median intra-aptamer correlation (SomaScan)",
    y = "Aptamer-specific Olink correlation",
    title = "Aptamer consistency vs Olink concordance"
  ) +
  theme_bw()

ggsave("results/aptamer_intra_vs_olink_scatter.png", width = 6, height = 4)
```

Correlation vs abundance and variance

```{r}
# Olink stats per Assay (all NPX observations)
olink_stats <- olink |>
  group_by(Assay) |>
  summarise(
    mean_olink = mean(NPX, na.rm = TRUE),
    median_olink = median(NPX, na.rm = TRUE),
    var_olink = var(NPX, na.rm = TRUE),
    sd_olink = sd(NPX, na.rm = TRUE),
    cv_olink = ifelse(mean_olink != 0, sd_olink/mean_olink, NA_real_),
    .groups = "drop"
  )

# Soma stats per AptName (Value)
soma_stats <- soma |>
  group_by(AptName) |>
  summarise(
    mean_soma = mean(Value, na.rm = TRUE),
    median_soma = median(Value, na.rm = TRUE),
    var_soma = var(Value, na.rm = TRUE),
    sd_soma = sd(Value, na.rm = TRUE),
    cv_soma = ifelse(mean_soma != 0, sd_soma/mean_soma, NA_real_),
    .groups = "drop"
  )

# Join stats to comparison table
cmp_stats <- comparison |>
  left_join(olink_stats, by = "Assay") |>
  left_join(soma_stats, by = "AptName") |>
  # avoid zeros for log CV
  mutate(logCV_olink = log10(cv_olink + 1),
         logCV_soma  = log10(cv_soma  + 1)) |>
  filter(!is.na(r_all))

# Olink abundance/variance vs correlation
ggplot(cmp_stats, aes(x = mean_olink, y = r_all, , color = bin)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "Mean Olink (NPX)",
    y = "Spearman correlation (all samples)"
  ) +
  scale_colour_brewer(palette = "Set1") +
  theme_bw() 

ggsave("results/mean_olink_vs_corr.png", width = 6, height = 5)

ggplot(cmp_stats, aes(x = logCV_olink, y = r_all, , color = bin)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "Log10(CV) Olink (NPX)",
    y = "Spearman correlation (all samples)"
  ) +
  scale_colour_brewer(palette = "Set1") +
  theme_bw() 

ggsave("results/cv_olink_vs_corr.png", width = 6, height = 5)

# Soma abundance/variance vs correlation
ggplot(cmp_stats, aes(x = mean_soma, y = r_all, , color = bin)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "Mean SomaScan (Value)",
    y = "Spearman correlation (all samples)"
  ) +
  scale_colour_brewer(palette = "Set1") +
  theme_bw() 

ggsave("results/mean_soma_vs_corr.png", width = 6, height = 5)

ggplot(cmp_stats, aes(x = logCV_soma, y = r_all, , color = bin)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "Log10(CV) SomaScan (Value)",
    y = "Spearman correlation (all samples)"
  ) +
  scale_colour_brewer(palette = "Set1") +
  theme_bw() 

ggsave("results/cv_soma_vs_corr.png", width = 6, height = 5)

# Mean/CV olink vs Mean/CV Soma vs Correlation
ggplot(cmp_stats, aes(x = mean_olink, y = mean_soma, z = r_all)) +
  stat_summary_2d(fun = mean, bins = 50, na.rm = TRUE) +
  scale_fill_viridis_c(option = "B", na.value = "grey50") +
  labs(x = "Mean Olink (NPX)", y = "Mean SomaScan (Value)", fill = "Mean r") +
  theme_bw()

ggsave("results/mean_soma_vs_mean_olink_heatmap.png", width = 6, height = 5)

ggplot(cmp_stats, aes(x = logCV_olink, y = logCV_soma, z = r_all)) +
  stat_summary_2d(fun = mean, bins = 50, na.rm = TRUE) +
  scale_fill_viridis_c(option = "B", na.value = "grey50") +
  labs(x = "Log10(CV) Olink (NPX)", y = "Log10(CV) SomaScan (Value)", fill = "Mean r") +
  theme_bw()

ggsave("results/cv_soma_vs_cv_olink_heatmap.png", width = 6, height = 5)
```

Deming regression per assay/aptamer. Most proteins show small proportional and additive biases: slopes near 0–1, intercepts ~10.
A few extreme outliers distort summary statistics; these likely correspond to very low-abundance proteins or proteins with multiple aptamers that are inconsistent.

```{r}
merged_data <- olink |>
  inner_join(soma, by = c("DAid", "Assay"))

# Compute Deming regression per protein
groups <- merged_data |> group_by(Assay, AptName) |> group_split()
group_keys <- merged_data |> group_by(Assay, AptName) |> group_keys()

plan(multisession, workers = 8)
handlers(global = TRUE)

deming_results <- future_map2_dfr(groups, seq_along(groups), ~{
  df <- .x
  ok <- is.finite(df$NPX) & is.finite(df$Value)
  if(sum(ok) > 2){
    fit <- mcreg(x = df$NPX[ok], y = df$Value[ok], method.reg = "Deming", error.ratio = 1)
    tibble(Assay = df$Assay[1], AptName = df$AptName[1],
           slope = fit@para[2,1], intercept = fit@para[1,1])
  } else {
    tibble(Assay = df$Assay[1], AptName = df$AptName[1],
           slope = NA_real_, intercept = NA_real_)
  }
}, .progress = TRUE, seed = TRUE)

# Filter out extreme outliers
slope_limits <- quantile(deming_results$slope, probs = c(0.005, 0.995), na.rm = TRUE)
intercept_limits <- quantile(deming_results$intercept, probs = c(0.005, 0.995), na.rm = TRUE)

deming_filtered <- deming_results |>
  filter(slope >= slope_limits[1], slope <= slope_limits[2],
         intercept >= intercept_limits[1], intercept <= intercept_limits[2])

# Plot slope distribution
ggplot(deming_filtered, aes(x = slope)) +
  geom_histogram(binwidth = 0.1, fill = "lightblue", color = "black") +
  labs(title = "Deming slopes (SomaScan ~ Olink)", x = "Slope", y = "Count") +
  theme_bw()

ggsave("results/deming_slope_distribution.png", width = 5, height = 4)

# Plot intercept distribution
ggplot(deming_filtered, aes(x = intercept)) +
  geom_histogram(binwidth = 0.1, fill = "salmon", color = "black") +
  labs(title = "Deming intercepts (SomaScan ~ Olink)", x = "Intercept", y = "Count") +
  theme_bw()

ggsave("results/deming_intercept_distribution.png", width = 5, height = 4)

# Plot slope vs mean abundance
mean_abundance <- merged_data |>
  group_by(Assay, AptName) |>
  summarise(mean_olink = mean(NPX, na.rm = TRUE),
            mean_soma  = mean(Value, na.rm = TRUE),
            .groups = "drop")

deming_plot <- deming_filtered |>
  left_join(mean_abundance, by = c("Assay", "AptName"))

ggplot(deming_plot, aes(x = mean_olink, y = slope)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  labs(x = "Mean Olink NPX", y = "Deming slope", title = "Proportional bias vs mean abundance") +
  theme_bw()

ggsave("results/deming_slope_vs_mean_olink.png", width = 5, height = 4)
```

## Plot individual assay scatter plots

Example low correlation proteins as scatterplot (Olink vs Soma)

```{r}
random_faceted_plot <- function(correlation, olink, soma, n, threshold = 0.5) {
  assays_to_select <- correlation |>
    filter(correlation < threshold)
  
  if (nrow(assays_to_select) <= 1) {
    stop("Not enough correlations with correlation lower than 0.5 to select.")
  }

  # Randomly select n Assay values from the filtered data frame
  selected_assays <- assays_to_select |>
    dplyr::slice_sample(n = n) |>
    pull(Assay)

  # Select the corresponding rows from the soma and olink data frames
  selected_dat <- soma |>
    filter(Assay %in% selected_assays) |>
    left_join(olink |> filter(Assay %in% selected_assays), by = c("DAid", "Assay")) |>
    left_join(correlation |> select(Assay, AptName, correlation), by = c("Assay", "AptName"))
  
  label_df <- selected_dat |>
    group_by(Assay, AptName, correlation) |>
    summarize(x = max(NPX, na.rm = TRUE),
              y = max(Value, na.rm = TRUE),
              .groups = "drop")

  # Create the faceted plot
  ggplot(selected_dat, aes(x = NPX, y = Value)) +
    geom_point() +
    geom_text(data = label_df,
              aes(x = Inf, y = Inf, label = paste0("r = ", round(correlation, 3))),
              hjust = 1, vjust = 1, size = 3) +
    facet_wrap(~Assay + AptName) +
    labs(x = "Olink NPX", y = "Soma Value") +
    theme_bw()
}

random_faceted_plot(correlation, olink, soma, n = 10, threshold = 0)
```

Plot selected assays

```{r}
selected_assays <- c("DDC", "VWF", "NANOG", "NPPB", "SMCO3", "CILP", "FN1", "CHUK", "TRAF2", "CNDP1", "DEFA6", "TUBA1A", "SUMF1")

above_lod <- olink |>
  group_by(Assay) |>
  summarise(perc_above_lod = mean(NPX >= LOD, na.rm = TRUE) * 100)

selected_dat <- soma |>
  filter(Assay %in% selected_assays) |>
  left_join(olink |> filter(Assay %in% selected_assays), by = c("DAid", "Assay")) |>
  left_join(correlation |> select(Assay, AptName, correlation), by = c("Assay", "AptName")) |>
  left_join(above_lod, by = "Assay")
  
label_df <- selected_dat |>
  group_by(Assay, AptName, correlation, perc_above_lod) |>
  summarise(x = max(NPX, na.rm = TRUE),
            y = max(Value, na.rm = TRUE),
            .groups = "drop") |>
  mutate(label = paste0(
    "r = ", round(correlation, 3), 
    "\nAbove LOD = ", round(perc_above_lod, 1), "%"
  ))

  # Create the faceted plot
ggplot(selected_dat, aes(x = NPX, y = Value)) +
  geom_point() +
  geom_text(data = label_df,
          aes(x = Inf, y = Inf, label = label),
          hjust = 1.05, vjust = 1.2, size = 3) +
  facet_wrap(~Assay + AptName) +
  labs(x = "Olink NPX", y = "Soma Value") +
  theme_bw()
```